### 2018-01-23 submission/manuscript.tex
fix abstract and other grammar errors

\textbf{Conclusions.} This data provides support for the importance of
the bacterial community to both adenoma and carcinoma genesis. The
evidence collected within this study on the role of the microbiota in
CRC pathogenesis shows stronger associations between carcinoma then
adenoma. One reason for this may be in part due to the low power to
detect more subtle changes in the majority of studies that have been
performed to date.

@@

Colorectal cancer (CRC) is a growing world-wide health problem {[}1{]}
in which the microbiota has been purported to play an active role in
disease pathogenesis {[}2{]}. Numerous studies have shown the importance
of both individual microbes {[}3--7{]} and the overall community
{[}8--10{]} in tumorgenesis using mouse models of CRC. There have also
been numerous case/control studies investigating the microbiota in the
formation of both adenoma and carcinoma. A recent meta-analysis
investigated whether specific biomarkers could be consistently
identified using multiple data sets {[}11{]}. This meta-analysis focused
on identifying biomarkers or individual microbes but did not critically
investigate how the community changes in CRC.

Although there has been an intense focus on microbiota based biomarker
discovery for CRC, the candidate genera seem to be numerous. Some
studies point towards mouth-associated genera such as
\emph{Fusobacterium}, \emph{Peptostreptococcus}, \emph{Parvimonas}, and
\emph{Porphyromonas} as key enriched genera {[}6,12--18{]}. Yet, even in
some of these studies mouth-associated genera are far from the only
microbes identified to be associated with CRC. These other genera
include, but are not necessarily limited to, \emph{Providencia},
\emph{Mogibacterium}, \emph{Enterococcus}, \emph{Escherichia/Shigella},
\emph{Klebsiella}, and \emph{Streptococcus} {[}14--16{]}. In fact,
\emph{Escherichia/Shigella} and \emph{Streptococcus} also have good
\emph{in vivo} evidence that individual species within the genus can be
important in the pathogenesis of CRC {[}5,19,20{]}. Other studies have
also identified \emph{Akkermansia muciniphila} and \emph{Bacteroides
fragilis} as potential additional markers of CRC with \emph{Bacteroides
fragilis} having good mechanistic studies into how this may occur
{[}15,21,22{]}. Still, other studies have identified genera like
\emph{Roseburia} as increased in CRC while others have reported it to be
decreased or have found no difference in any genera at all
{[}15,18,23,24{]}.

@@

Targeting the identification of biomarkers within stool seems logical
since it offers an easy and cost-effective way to stratify risk and the
current gold standard for diagnosis, a colonoscopy, can be
time-consuming and is not risk free. Although stool represents an easy
and less invasive way to assess risk, it is not clear how reflective
this sample is to the community on the adenoma or carcinoma. Some
studies have tried to assess this in health and disease but are limited
by their sample size {[}18,28{]}. Sampling the tissue directly may
provide clearer answers but is not without limitations. Due to bowel
prep the communities left for sampling may not be reflective of the
resident microbiota, but rather a collection of what is able to keep
adhered to the mucosa. Additionally, these samples contain more host
DNA, potentially limiting the types of analysis that can be done. It is
well known that low biomass samples can be very difficult to work with
and results can end up being study dependent due to the randomness of
contamination {[}29{]}. Due to these many differences that could arise
between stool and tissue, one question our meta-analysis aims to answer
is whether there are consistent patterns that emerge across studies
regardless of whether they used stool or tissue samples.

@@

Using both feces (n = 1737) and tissue (492 samples from 350
individuals) totaling over 2229 total samples across 14 studies
{[}12--18,21,23--27,30{]} {[}Table 1 \& 2{]}, we expand both the breadth
and scope of the previous meta-analysis to investigate whether the
bacterial community is an important risk factor or if select members can
recapitulate the same information for both adenoma and carcinoma. To
accomplish this we first assessed whether bacterial diversity changes
throughout disease (control to adenoma to carcinoma) and if it results
in an increased relative risk (RR) for adenoma or carcinoma. We then
assessed what genera, if any, increase or decrease the RR of adenoma or
carcinoma. Next, using Random Forest models, we analyzed whether the
full community or only the top 5 increased and top 5 decreased RR genera
combined resulted in better model classification, based on the area
under the curve (AUC). Our results suggest that the community changes as
disease severity worsens and that this community is important for
disease classification. Finally, we also examined at what effect and
sample size the studies that were used were powered for and the sample
size needed to get to an 80\% power. Our power analysis suggests that
each individual study used was underpowered for detecting effect size
differences of 10\% or below between the case and control groups.

@@

\textbf{\emph{Lower Bacterial Diversity is Associated with Increased RR
of Carcinomas:}} Power transforming and Z-score normalizing OTU
richness, evenness, and Shannon diversity for the entire data set
allowed for the combination of data from all studies to assess
differences in broad scale community metrics as disease severity
worsens. Using linear mixed-effect models to control for study and
variable region we assessed whether OTU richness, evenness, or Shannon
diversity changed in a step-wise manner with disease severity. In stool,
there was a significant decrease in both evenness and Shannon diversity
from control to adenoma to carcinoma (P-value = 0.025 and 0.043,
respectively) {[}Figure 1A{]}. We next tested whether these detectable
differences in community resulted in significant increases in RR of
lesion (adenoma or carcinoma). For fecal samples, a decrease versus the
overal median in evenness resulted in a significantly increased RR for
carcinoma (RR = 1.36 (1.15 - 1.61), P-value = 0.00034) and adenoma (RR =
1.16 (1.01 - 1.34), P-value = 0.032) while a decrease versus the overal
median in Shannon diversity only increased the RR for carcinoma (RR =
1.33 (1.09 - 1.62), P-value = 0.0047) {[}Figure 2{]}. Using the
Bray-Curtis distance metric and PERMANOVA, it was also possible to
identify significant bacterial community changes, in specific studies,
for both carcinoma versus control and adenoma versus control {[}Table S1
\& S2{]}.

@@

\textbf{\emph{Mouth-Associated Genera are Associated with an Increased
RR of Lesion:}} Based on the small increase in RR using
\(\alpha\)-diversity metrics, we assessed if being higher than the
median for any specific genera resulted in an altered RR for adenoma or
carcinoma in stool and tissue. To investigate this we analyzed all
common genera across each study, in tissue or stool, and assessed
whether a relative abundance higher than the median results in an
increase or decrease in RR. For both stool and tissue none of the genera
in the top 5 most significantly increased RR were the same between
adenoma and carcinoma {[}Figure 3{]}. Mouth-associated genera were in
the top 5 genera associated with an increased RR of adenoma
(\emph{Pyramidobacter} {[}Figure 3A{]} and \emph{Rothia} {[}Figure
3C{]}) and carcinoma (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Porphyromonas}, and \emph{Peptostreptococcus} {[}Figure 3B{]} and
\emph{Fusobacterium} {[}Figure 3D{]}) for both stool and tissue.
Conversely, genera commonly associated with a normal gastrointestinal
tract were correlated with a decreased RR for both adenoma and carcinoma
for both stool and tissue {[}Figure 3{]}. Overall, there was little
direct overlap of the top 5 increased or decreased RR genera between
stool and tissue.

@@

\textbf{\emph{Whole Community Models Add Important Context:}} Since
specific genera increased RR for carcinoma over diversity metrics we
assessed whether the bacterial community was better at classifying
disease versus only a select group of genera based on RR and P-value
significance and tested this using two approaches. The first approach
used genus level data and tested whether there were any differences in
AUC between all genera and selected genera during training on one study
or testing on all other studies. The second approach used OTU level data
and tested whether there was a generalized decrease in the 10-fold cross
validation (CV) model AUC across studies when comparing all OTUs versus
only OTUs that taxonomically classified to selected genera.

@@

In stool the most common genera in the full genera-based models belonged
to genera such as \emph{Ruminococcus}, \emph{Bacteroides}, and
\emph{Roseburia} {[}Figure 6A \& B{]}. Although all four CRC-associated
genera were present in carcinoma, none were present in the majority of
studies and \emph{Fusobacterium} was the only genus present in adenoma
{[}Figure 6A \& B{]}. For the full OTU-based models,
\emph{Ruminococcaceae} was present in the top 10 consistently for both
adenoma and carcinoma models while \emph{Roseburia} was only present in
many adenoma models and \emph{Bacteroides} was present in the
overwhelming majority of the carcinoma models {[}Figure 6C \& 6D{]}.

@@

Our study identifies clear differences in diversity, both at the
community level and for individual genera, that are present in patients
with and without CRC {[}Figure 1-3{]}. Although there was a step-wise
decrease in diversity from control to adenoma to carcinoma, this did not
translate into large effect sizes for the RR of lesion. Even though
mouth-associated genera increased the RR of carcinoma, they did not
consistently increase the RR of adenoma. These mouth-associated genera
are clearly important to carcinoma classification but our observations
suggest that accounting for the community in which these microbes exist
can increase the ability of models to make predictions.

The data presented herein supports the importance of select genera for
carcinoma but not necessarily adenoma formation. Our observations show
that when only using the select genera to create models for carcinoma
the AUC are similar to using all genera or OTUs {[}Figure 4 \& 5{]}.
This suggests that an interplay between a select number of potentially
protective and exacerbating microbes, within the GI community, are
crucial for carcinoma formation. Importantly, it suggests that there may
be key members of the community that might be studied further to
potentially reduce the risk of carcinoma. Conversely, using the present
data, it is clear that new approaches may need to be used to identify
members of the community that are associated with adenoma. Regardless of
sample type and whether a full or select model was used our Random
Forest models consistently performed poorly. Yet, the step-wise decrease
in diversity suggests that the adenoma community is not normal but has
changed subtly {[}Figure 1{]}. This change in diversity, at this early
stage of disease, could be focal to the adenoma itself. One potential
hypothesis from these observations is that at early stages of the
disease, how the host interacts with these subtle changes is what
ultimately leads to a thoroughly dysfunctional community that is
supportive of CRC genesis.

@@

The tissue studies did not provide a clearer understanding of how the
microbiota may be associated with lesion. Generally, the full OTU-based
models of unmatched and matched {[}Figure S7E \& F{]} tissue samples
were concordant with stool showing that resident microbes were the most
prevalent in the top 10 across study. Unlike in stool,
\emph{Fusobacterium} was the only mouth-associated bacteria consistently
present in the top 10 of the full CRC models {[}Figure S7B-C \& E-F{]}.
The majority of the tissue results seem to be study specific with many
top 10 taxa being present only in a single study. Additionally, the
presence of genera associated with contamination, within the top 10 most
important variables for the genera and OTU models, is worrying because
it is commonly a marker of contamination. The low bacterial biomass of
tissue samples coupled with potential contamination could ultimately
explain why the these results seem to be more sporadic than the stool
results.

One important caveat to this study is that even though specific genera
associated with species such as \emph{Bacteroides fragilis} and
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} were not
identified, it does not necessarily mean these specific species are not
important in human CRC {[}20,22{]}. Since we are limited in our
aggregation of the data, across study, to the genus level, it is not
possible to clearly delineate what specific species are contributing to
overall disease progression. Our observations are not inconsistent with
the previous literature on either \emph{Bacteroides fragilis} or
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus}. As an
example, for stool, the full RF models consistently identified the genus
\emph{Bacteroides} as well as OTUs that classified as \emph{Bacteroides}
to be important model component across studies. Suggesting that even
though \emph{Bacteroides} may not increase the RR of lesion and may not
vary in relative abundance, like \emph{Fusobacterium}, it is still
important in CRC. Additionaly, \emph{Streptococcus gallolyticus} subsp.
\emph{gallolyticus} is a mouth-associated micobe and our study clearly
shows that regardless of whether the sample is tissue or stool,
mouth-associated genera are commonly associated with an increased RR of
both adenoma and carcinoma.

The associations between the microbiota and adenoma are inconclusive, in
part, because many studies may not be powered effectively to observe
small effect sizes. None of the studies analyzed were properly powered
to detect a 10\% or lower change between cases and controls. A small
effect size may well be the scope in which differences consistently
occur in adenoma based on the results observed within our meta-analysis.
Future studies investigating adenoma and the microbiota need to take
power into consideration if we are to reproducibly study whether the
microbiota contributes to poly formation. In contrast to adenoma, our
observations suggest that most studies analyzed have sufficient power to
detect many changes in carcinoma because of large effect size
differences between cases and controls {[}Figure 7{]}.

>>>>

\textbf{Conclusions.} This data provides support for the importance of
the bacterial community to both adenoma and carcinoma genesis. The
evidence collected within this study on the role of the microbiota in
CRC pathogenesis shows stronger associations between carcinoma then
adenoma and this may be in part due to the low power to detect more
subtle changes in the majority of studies that have been performed to
date.

@@

Colorectal cancer (CRC) is a growing world-wide health problem in which
the microbiota has been purported to play an active role in disease
pathogenesis {[}1,2{]}. Numerous studies have shown the importance of
both individual microbes {[}3--7{]} and the overall community
{[}8--10{]} in tumorgenesis using mouse models of CRC. There have also
been numerous case/control studies investigating the microbiota in the
formation of both adenoma and carcinoma. A recent meta-analysis
investigated whether specific biomarkers could be consistently
identified using multiple data sets {[}11{]}. This meta-analysis focused
on identifying biomarkers or individual microbes but did so on a small
total number of individuals and only investigated stool.

Although there has been an intense focus on microbiota-based biomarker
discovery for CRC, the candidate genera seem to be numerous. Some
studies point towards mouth-associated genera such as
\emph{Fusobacterium}, \emph{Peptostreptococcus}, \emph{Parvimonas}, and
\emph{Porphyromonas} as key enriched genera {[}6,12--18{]}. Yet, even in
some of these studies mouth-associated genera are far from the only
microbes identified to be associated with CRC. These other genera
include, but are not necessarily limited to, \emph{Providencia},
\emph{Mogibacterium}, \emph{Enterococcus}, \emph{Escherichia/Shigella},
\emph{Klebsiella}, and \emph{Streptococcus} {[}14--16{]}. In fact,
\emph{Escherichia/Shigella} and \emph{Streptococcus} also have good
\emph{in vivo} evidence that individual species within the genus can be
important in the pathogenesis of CRC {[}5,19,20{]}. Other studies have
also identified \emph{Akkermansia muciniphila} and \emph{Bacteroides
fragilis} as potential additional markers of CRC with \emph{Bacteroides
fragilis} having good mechanistic studies into how this may occur
{[}15,21,22{]}. Still, other studies have identified genera like
\emph{Roseburia} as increased in CRC while others have reported it to be
decreased or have found no difference in any genera at all
{[}15,18,23,24{]}.

@@

Targeting the identification of biomarkers within stool seems logical
since it offers an easy and cost-effective way to stratify risk and the
current gold standard for diagnosis, a colonoscopy, can be
time-consuming and is not risk free. Although stool represents an easy
and less invasive way to assess risk, it is not clear how reflective
this sample is to the community on the adenoma or carcinoma. Some
studies have tried to assess this in health and disease but are limited
by their sample size {[}18,28{]}. Sampling the tissue directly may
provide clearer answers but is not without limitations. Due to bowel
prep the communities left for sampling may not be reflective of the
resident microbiota, but rather a collection of what is able to keep
adhered to the mucosa. Additionally, these samples contain more host
DNA, potentially limiting the types of analysis that can be done. It is
well known that low biomass samples can be very difficult to work with
and results can end up being study dependent due to the randomness of
contamination {[}29{]}. Due to these many differences that could arise
between stool and tissue, an additional question our meta-analysis aims
to answer is whether there are consistent patterns that emerge across
studies regardless of whether they used stool or tissue samples.

@@

Using both feces (n = 1737) and tissue (492 samples from 350
individuals) totaling over 2229 total samples across 14 studies
{[}12--18,21,23--27,30{]} {[}Table 1 \& 2{]}, we expand both the breadth
and scope of the previous meta-analysis to investigate whether the
bacterial community is an important risk factor or if select members can
recapitulate the same information for both adenoma and carcinoma. To
accomplish this we first assessed whether bacterial diversity changes
throughout disease (control to adenoma to carcinoma) and if it results
in an increased relative risk (RR) for adenoma or carcinoma. We then
assessed what genera, if any, increase or decrease the RR of adenoma or
carcinoma. Next, using Random Forest models, we analyzed whether the
full community or only the top 5 increased and top 5 decreased RR genera
combined resulted in better model classification, based on the area
under the curve (AUC). Our results suggest that the community changes as
disease severity worsens and that this community is important for
disease classification. Finally, we also examined at what effect and
sample size the studies that were used were powered for and the sample
size needed to get to an 80\% power.

@@

\textbf{\emph{Lower Bacterial Diversity is Associated with Increased RR
of Carcinomas:}} Power transforming and Z-score normalizing Operational
Taxonomic Unit (OTU) richness, evenness, and Shannon diversity for the
entire data set allowed for the combination of data from all studies to
assess differences in broad scale community metrics as disease severity
worsens. Using linear mixed-effect models to control for study and
variable region we assessed whether OTU richness, evenness, or Shannon
diversity changed in a step-wise manner with disease severity. In stool,
there was a significant decrease in both evenness and Shannon diversity
from control to adenoma to carcinoma (P-value = 0.025 and 0.043,
respectively) {[}Figure 1A{]}. We next tested whether these detectable
differences in community resulted in significant increases in RR of
lesion (adenoma or carcinoma). For fecal samples, a decrease versus the
overal median in evenness resulted in a significantly increased RR for
carcinoma (RR = 1.36 (1.15 - 1.61), P-value = 0.00034) and adenoma (RR =
1.16 (1.01 - 1.34), P-value = 0.032) while a decrease versus the overal
median in Shannon diversity only increased the RR for carcinoma (RR =
1.33 (1.09 - 1.62), P-value = 0.0047) {[}Figure 2{]}. Using the
Bray-Curtis distance metric and PERMANOVA, it was also possible to
identify significant bacterial community changes, in specific studies,
for both carcinoma versus control and adenoma versus control {[}Table S1
\& S2{]}.

@@

\textbf{\emph{Mouth-Associated Genera are Associated with an Increased
RR of Lesion:}} Based on the small increase in RR using OTU richness and
Shannon diversity, we assessed if being higher than the median for any
specific genera resulted in an altered RR for adenoma or carcinoma in
stool and tissue. To investigate this we analyzed all common genera
across each study, in tissue or stool, and assessed whether a relative
abundance higher than the median results in an increase or decrease in
RR. For both stool and tissue none of the genera in the top 5 most
significantly increased RR were the same between adenoma and carcinoma
{[}Figure 3{]}. Mouth-associated genera were in the top 5 genera
associated with an increased RR of adenoma (\emph{Pyramidobacter}
{[}Figure 3A{]} and \emph{Rothia} {[}Figure 3C{]}) and carcinoma
(\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Porphyromonas}, and
\emph{Peptostreptococcus} {[}Figure 3B{]} and \emph{Fusobacterium}
{[}Figure 3D{]}) for both stool and tissue. Conversely, genera commonly
associated with a normal gastrointestinal tract were correlated with a
decreased RR for both adenoma and carcinoma for both stool and tissue
{[}Figure 3{]}. Overall, there was little direct overlap of the top 5
increased or decreased RR genera between stool and tissue.

@@

\textbf{\emph{Select Genera Models can Recapitulate Whole Community
Models:}} Since specific genera increased RR for carcinoma over
diversity metrics we assessed whether the bacterial community was better
at classifying disease versus only a select group of genera based on RR
and P-value significance and tested this using two approaches. The first
approach used genus level data and tested whether there were any
differences in AUC between all genera and selected genera during
training on one study or testing on all other studies. The second
approach used OTU level data and tested whether there was a generalized
decrease in the 10-fold cross validation (CV) model AUC across studies
when comparing all OTUs versus only OTUs that taxonomically classified
to selected genera.

@@

In stool, the most common genera in the full genera-based models
belonged to genera such as \emph{Ruminococcus}, \emph{Bacteroides}, and
\emph{Roseburia} {[}Figure 6A \& B{]}. Although all mouth-associated
genera were present in carcinoma, none were present in the majority of
studies and \emph{Fusobacterium} was the only genus present in adenoma
{[}Figure 6A \& B{]}. For the full OTU-based models,
\emph{Ruminococcaceae} was present in the top 10 consistently for both
adenoma and carcinoma models while \emph{Roseburia} was only present in
many adenoma models and \emph{Bacteroides} was present in the
overwhelming majority of the carcinoma models {[}Figure 6C \& 6D{]}.

@@

Our study identifies clear differences in diversity, both at the
community level and for individual genera, that are present in patients
with and without CRC {[}Figure 1-3{]}. Although there was a step-wise
decrease in diversity from control to adenoma to carcinoma, this did not
translate into large effect sizes for the RR of lesion. Even though
mouth-associated genera increased the RR of carcinoma, they did not
consistently increase the RR of adenoma. These mouth-associated genera
are clearly important to carcinoma classification but our observations
suggest that accounting for other community members can increase the
ability of models to make predictions.

The data presented herein supports the importance of select genera for
carcinoma but not necessarily adenoma formation. Our observations show
that when only using the select genera to create models for carcinoma
the AUC are similar to using all genera or OTUs {[}Figure 4 \& 5{]}.
This suggests that an interplay between a select number of potentially
protective and exacerbating microbes, within the GI community, are
crucial for carcinoma formation. Importantly, it suggests that there may
be key members of the GI community that might be studied further to
potentially reduce the risk of carcinoma. Conversely, using the present
data, it is clear that new approaches may need to be used to identify
members of the community that are associated with adenoma. Regardless of
sample type and whether a full or select model was used our Random
Forest models consistently performed poorly. Yet, the step-wise decrease
in diversity suggests that the adenoma community is not normal but has
changed subtly {[}Figure 1{]}. This change in diversity, at this early
stage of disease, could be focal to the adenoma itself. One potential
hypothesis from these observations is that at early stages of the
disease, how the host interacts with these subtle changes is what
ultimately leads to a thoroughly dysfunctional community that is
supportive of CRC genesis.

@@

The tissue studies did not provide a clearer understanding of how the
microbiota may be associated with lesion. Generally, the full OTU-based
models of unmatched and matched {[}Figure S7E \& F{]} tissue samples
were concordant with stool showing that resident microbes were the most
prevalent in the top 10 across study. Unlike in stool,
\emph{Fusobacterium} was the only mouth-associated bacteria consistently
present in the top 10 of the full CRC models {[}Figure S7B-C \& E-F{]}.
The majority of the tissue results seem to be study specific with many
top 10 taxa being present only in a single study. Additionally, the
presence of genera associated with contamination, within the top 10 most
important variables for the genera and OTU models is worrying. The low
bacterial biomass of tissue samples coupled with potential contamination
could ultimately explain why the these results seem to be more sporadic
than the stool results.

One important caveat to this study is that even though specific genera
associated with species such as \emph{Bacteroides fragilis} and
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} were not
identified, it does not necessarily mean these specific species are not
important in human CRC {[}20,22{]}. Since we are limited in our
aggregation of the data, across study, to the genus level, it is not
possible to clearly delineate what specific species are contributing to
overall disease progression. Our observations are not inconsistent with
the previous literature on either \emph{Bacteroides fragilis} or
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus}. As an
example, for stool, the full RF models consistently identified the genus
\emph{Bacteroides} as well as OTUs that classified as \emph{Bacteroides}
to be important model components across studies. Suggesting that even
though \emph{Bacteroides} may not increase the RR of lesion and may not
vary in relative abundance, like \emph{Fusobacterium}, it is still
important in CRC. Additionaly, \emph{Streptococcus gallolyticus} subsp.
\emph{gallolyticus} is a mouth-associated micobe and our study clearly
shows that regardless of whether the sample is tissue or stool,
mouth-associated genera are commonly associated with an increased RR of
both adenoma and carcinoma.

The associations between the microbiota and adenoma are inconclusive, in
part, because many studies may not be powered effectively to observe
small effect sizes. None of the studies analyzed were properly powered
to detect a 10\% or lower change between cases and controls. A small
effect size may well be the scope in which differences consistently
occur in adenoma based on the results observed within our meta-analysis.
Future studies investigating adenoma and the microbiota need to take
power into consideration if we are to reproducibly study whether the
microbiota contributes to polyp formation. In contrast to adenoma, our
observations suggest that most studies analyzed have sufficient power to
detect many changes in carcinoma because of large effect size
differences between cases and controls {[}Figure 7{]}.
